---
title: "Making plots and tables from the simulation results"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
source_rmd <- function(file, local = FALSE){
  options(knitr.duplicate.label = "allow")
  tempR <- tempfile(tmpdir = ".", fileext = ".R")
  on.exit(unlink(tempR))
  print(file.exists("model_functions.Rmd"))
  knitr::purl(file, output = tempR, quiet = TRUE)
  source(tempR, local = globalenv())
}
source_rmd("analysis/model_functions.Rmd") 
```

## Load packages and results

```{r load_data, results='hide', message=FALSE, warning=FALSE}
packages <- c("dplyr", "purrr", "ggplot2", "reshape2", "Cairo", "knitr", "stringr",
              "latex2exp", "pander", "grid", "gridExtra", "ggthemes", "data.table",
              "readr", "tibble", "biglm", "kableExtra", "future", "future.apply", "parallel")
lapply(packages, library, character.only = TRUE, quietly = TRUE)

if(!file.exists("data/all_results.rds")){
  
  files <- list.files(path = file.path(getwd(), "data"), pattern = "results_", full.names = TRUE)
  files <- files[str_detect(files, "results_[:digit:]+.rds")]
  n_files <- length(files)
  
  # Open all the results files and bind them together into a tibble
  results <- lapply(1:length(files), 
                    function(i) {
                      if(i %% 10 == 0) print(paste("Doing file", i, "of", n_files))
                      readRDS(files[i]) %>% filter(cost_A == 0 & cost_B == 0)}) %>% 
    rbindlist() %>% as_tibble()
  
  # Make sure each unique parameter space is run exactly once,
  # unless it was expressly run multiple times using "realisations" (never used in this paper)
  results <- results %>% 
    distinct(release_strategy, W_shredding_rate, Z_conversion_rate,
             Zr_creation_rate, Zr_mutation_rate, Wr_mutation_rate,
             cost_Zdrive_female, cost_Zdrive_male, male_migration_prob,
             female_migration_prob, migration_type, n_patches, softness,
             male_weighting, density_dependence_shape, max_fecundity, realisations,
             initial_A, initial_B, .keep_all = TRUE)
  
  saveRDS(results, file = "data/all_results.rds")
} else results <- read_rds("data/all_results.rds")

results$went_extinct <- ifelse(results$outcome == "Population extinct", 1, 0)
results$migration_type[results$migration_type == "local"] <- "Local"
results$migration_type[results$migration_type == "global"] <- "Global"

# Command to upload the results file to Spartan - needed for Spartan to queue up all the parameter spaces not already finished
# scp /Users/lholman/Rprojects/W_shredder/data/all_results.rds lukeholman@spartan:/data/projects/punim0243/W_shredder/data/all_results.rds

# Helper function to get a list of all the model parameters that vary between runs
find_variable_parameters <- function(dat){
  dat %>% 
    select(-id, -realisations, -generation_extinct, -generation_Zd_extinct, 
           -generation_W_extinct, -generation_Zd_fixed, -outcome, -went_extinct, 
           -initial_Wr, -initial_Zr) %>%
    sapply(function(x) length(unique(x))) %>% 
    keep(~.x > 1) %>% names()
}
variable_parameters <- find_variable_parameters(results)
# combinations <- apply(combn(variable_parameters, 2), 2, paste0, collapse = " x ") # all 2-way combos of parameters

# Make a data frame to convert R-friendly names to figure-friendly names
nice_names <- data.frame(
  original = variable_parameters,
  new = gsub("_", " ", variable_parameters),
  stringsAsFactors = FALSE) %>%
  mutate(
    new = gsub("rel", "Rel", new),
    new = gsub("W shredding rate", "Gene drive in females ($p_{shred}$)", new),
    new = gsub("Z conversion rate", "Gene drive in males ($p_{conv}$)", new),
    new = gsub("Zr creation rate", "Rate of NHEJ ($p_{nhej}$)", new),
    new = gsub("Zr mutation rate", "Z-linked resistance ($\\mu{_Z}$)", new),
    new = gsub("Wr mutation rate", "W-linked resistance ($\\mu{_W}$)", new),
    new = gsub("cost Zdrive female", "Cost of Z* to females ($c_f$)", new),
    new = gsub("cost Zdrive male", "Cost of Z* to males ($c_m$)", new),
    new = gsub("male migration prob", "Male dispersal frequency", new),
    new = gsub("feMale dispersal frequency", "Female dispersal frequency (xf)", new),
    new = gsub("Male dispersal frequency", "Male dispersal frequency (xm)", new),
    new = gsub("xm", "$x_m$", new), 
    new = gsub("xf", "$x_f$", new),
    new = gsub("migration type", "Dispersal type", new),
    new = gsub("n patches", "Number of patches ($k$)", new),
    new = gsub("softness", "Importance of local density ($\\psi$)", new),
    new = gsub("male weighting", "Male effect on density ($\\delta$)", new),
    new = gsub("density dependence shape", "Shape of density dependence ($\\alpha$)", new),
    new = gsub("max fecundity", "Maximum fecundity ($r$)", new),
    new = gsub("initial A", "Initial freq. W-shredding resistance allele A", new),
    new = gsub("initial B", "Initial freq. gene conversion resistance allele B", new),
    new = as.character(TeX(new))) %>% 
  mutate(new = gsub("mu", "\\mu", new))

# Split the results by the type of gene drive:

# A W-shredder with imperfect shredding (for supplement)
W_shredder <- results %>% filter(cost_Zdrive_female != 1)

# A W-shredder with perfect shredding (main figures)
perfect_W_shredder <- results %>% filter(cost_Zdrive_female != 1 & W_shredding_rate == 1 & Zr_creation_rate != 0)

# A female-sterilising gene drive (not plotted, none of these runs show extinction, even for parameter space 4 where I tried to help extinciton to happen)
female_sterilising <- results %>% filter(cost_Zdrive_female == 1)

# Function to format numbers nicely
pretty <- function(x) prettyNum(x, big.mark = ",", scientific = FALSE)

rm(results)
```

## Number of simulation runs
I essentially ran the simulation until the figures ceased to look any different when I added extra random parameter spaces (generated by Latin hypercube sampling). The following shows the number of simualtion runs for a 

```{r count_runs}
n_runs <- tibble(
  `Type of run` = c("W-shredder with p-shred in the range 0-1",
                    "W-shredder with p-shred = 1",
                    "Female-sterilising gene drive with c_f = 1"),
  `Number of parameter spaces sampled` = pretty(c(nrow(W_shredder),
                                           nrow(perfect_W_shredder),
                                           nrow(female_sterilising)))
)

saveRDS(n_runs, file = "data/number_of_runs.rds")

n_runs %>% kable() %>% kable_styling()
```


## Make Figure 2: three individual runs

### Make the figure

```{r make_fig2, results='hide'}
ids <- c("10021351271896", "10158427238012", "10003485256727")   # 10258132461533

plot_run <- function(){
  
  get_data <- function(model_id, allele_freqs_list, result_df, label){
    
    df <- pluck_allele_freqs(model_id, allele_freqs_list) 
    
    alleles_to_plot <- group_by(df, allele) %>% 
      summarise(uniques = length(unique(frequency))) %>% 
      filter(uniques > 1) %>% # Don't plot alleles that stay at 0 whole time
      pull(allele)
    
    alleles_to_plot <- c(alleles_to_plot, "N")
    df <- df[df$allele %in% alleles_to_plot, ]
    df$frequency[df$allele == "N"] <- df$frequency[df$allele == "N"] /
      max(df$frequency[df$allele == "N"])
    
    paras <- result_df %>% mutate(id = as.character(id)) %>% filter(id == model_id)
    last_generation <- tail(df,12) %>% select(allele, frequency)
    #   print(paras); print(last_generation)
    
    df %>% mutate(facet = label,
                  allele = replace(allele, allele == "Zd", "Z*"),
                  allele = replace(allele, allele == "females", "Females"))
  }
  
  # x <- allele_freqs <- read_rds("data/results_1.rds")
  #  x %>% filter(outcome == "Population extinct" )   
  #  filter(x, outcome == "Zd fixed without extinction" & initial_A == 0 & initial_B == 0)$id
  allele_freqs <- read_rds("data/allele_freqs_1.rds")
  
  data1 <- get_data(ids[1], allele_freqs, W_shredder, "A. Z* invades, causing extinction") %>% 
    filter(generation >= 45)
  data1 <- data1 %>%
    rbind(data.frame(generation = max(data1$generation)+1, allele = "N", 
                     frequency = 0, facet = "A. Z* invades, causing extinction"))
  
  data2 <- get_data(ids[2], allele_freqs, W_shredder, "B. Z* invades but fails to cause extinction") %>% 
    filter(generation >= 35)
  data3 <- get_data(ids[3], allele_freqs, W_shredder, "C. Resistance to Z* prevents extinction")
  
  # Pantone colours:
  Marina <- "#4F84C4"
  Navy <- "#223A5E"
  Grenadine <- "#DA413D"
  Primrose <- "#EED4D9"
  Lilac <- "#774D8E"
  Bluebell <- "#93B4D7"
  cols <- c(Navy, Primrose, Grenadine, Marina, Lilac, "#CDA323", Bluebell)
  
  rbind(data1, data2, data3) %>%
    ggplot(aes(x = generation, y = frequency, colour = allele, group = allele)) + 
    geom_vline(xintercept = 50, linetype = 3, colour = "grey15", size = 0.9) + 
    geom_line(size = 0.9, alpha = 0.9) + 
    facet_wrap(~facet, scales = "free_x") +
    theme_hc() + 
    scale_colour_manual(values = cols, name = "") +
    theme(strip.text = element_text(hjust = 0), 
          strip.background = element_rect(fill = "grey90"),
          legend.position = "top",
          axis.ticks.y = element_blank()) + 
    xlab("Generation") + ylab("Frequency")
}

fig2 <- plot_run()
```

```{r fig.height=6, fig.width=10}
fig2
```
<br></br>
**Figure 2**: Three illustrative runs of the simulation, showing evolution in response to the release of males carrying a _W_-shredder at Generation 50 (dotted line). In panel A, the driving $Z^*$ allele fixed very quickly, causing population extinction through a shortage of females. In panel B, the $Z^*$ allele fixed but did not cause extinction, because carrier females continued to produce some daughters due to incomplete _W_-shredding. In panel C, the $Z^*$ allele invaded, which selected for the _W_-shredding resistance allele _A_, causing $Z^*$ to go extinct by removing its transmission advantage. The population size \textit{N} is shown as a fraction of its maximum value of 10,000. Table S1 lists the parameter spaces used for these three runs.

```{r echo=FALSE, results='hide'}
cairo_pdf(file = "figures/fig_2.pdf", width = 9, height = 4); fig2; dev.off()
```

### Make the accompanying table

```{r}
table_S1 <- t(rbind(
  data.frame(Panel = "A", W_shredder %>% filter(id == ids[1])),
  data.frame(Panel = "B", W_shredder %>% filter(id == ids[2])),
  data.frame(Panel = "C", W_shredder %>% filter(id == ids[3]))) %>%
  select(release_strategy, W_shredding_rate, 
         Z_conversion_rate, Zr_creation_rate, Zr_mutation_rate,
         Wr_mutation_rate, cost_Zdrive_female, cost_Zdrive_male,
         male_migration_prob, female_migration_prob,
         migration_type, n_patches, softness, male_weighting,
         density_dependence_shape, max_fecundity, initial_A, initial_B))

nice_names2 <- data.frame(original = variable_parameters,
                          new = gsub("_", " ", variable_parameters),
                          stringsAsFactors = FALSE) %>%
  mutate(
    new = gsub("rel", "Rel", new),
    new = gsub("W shredding rate", "Gene drive in females (p_shred)", new),
    new = gsub("Z conversion rate", "Gene drive in males (p_conv)", new),
    new = gsub("Zr creation rate", "Rate of NHEJ (p_nhej)", new),
    new = gsub("Zr mutation rate", "Z-linked resistance (μ_Z)", new),
    new = gsub("Wr mutation rate", "W-linked resistance (μ_W)", new),
    new = gsub("cost Zdrive female", "Cost of Z* to females (c_f)", new),
    new = gsub("cost Zdrive male", "Cost of Z* to males (c_m)", new),
    new = gsub("male migration prob", "Male dispersal frequency", new),
    new = gsub("feMale dispersal frequency", "Female dispersal frequency (x_f)", new),
    new = gsub("Male dispersal frequency", "Male dispersal frequency (x_m)", new),
    new = gsub("migration type", "Dispersal type", new),
    new = gsub("n patches", "Number of patches (k)", new),
    new = gsub("softness", "Importance of local density (ψ)", new),
    new = gsub("male weighting", "Male effect on density (𝛿)", new),
    new = gsub("density dependence shape", "Shape of density dependence (α)", new),
    new = gsub("max fecundity", "Maximum fecundity (r)", new),
    new = gsub("initial A", "Intial freq. W-shredding resistance allele A", new),
    new = gsub("initial B", "Intial freq. gene conversion resistance allele B", new)) 

rownames(table_S1) <- nice_names2$new[match(rownames(table_S1), nice_names2$original)]
rownames(table_S1) <- gsub("[.]", "", rownames(table_S1))
colnames(table_S1) <- c("Panel A", "Panel B", "Panel C")
table_S1[table_S1 == "one_patch"] <- "All in one patch"
table_S1[table_S1 == "all_patches"] <- "Scattered over all patches"
table_S1 <- as.data.frame(table_S1, stringsAsFactors = FALSE)

for(i in 1:nrow(table_S1)){
  if(!(table_S1[i, 1] %in% c("Local", "Scattered over all patches"))){
      table_S1[i, ] <- format(round(as.numeric(table_S1[i, ]), 2), nsmall = 2)
      if(max(as.numeric(table_S1[i, ])) > 5) table_S1[i, ] <- format(round(as.numeric(table_S1[i, ])), nsmall = 0)
}}
```

**Table S1**: List of the parameter values used to generate the simulation runs shown in Figure 2.
<br></br>
```{r}
table_S1 %>% kable(format = "html", escape = FALSE, digits=2) %>% kable_styling()
```

```{r echo=FALSE}
rownames(table_S1)[2]  <- "Gene drive in females ($p_{shred}$)"
rownames(table_S1)[3]  <- "Gene drive in males ($p_{conv}$)"
rownames(table_S1)[4]  <- "Rate of NHEJ ($p_{nhej}$)"
rownames(table_S1)[5]  <- "Z-linked resistance ($\\mu_Z$)"
rownames(table_S1)[6]  <- "W-linked resistance ($\\mu_W$)"
rownames(table_S1)[7]  <- "Cost of $Z^*$ to females ($c_f$)"
rownames(table_S1)[8]  <- "Cost of $Z^*$ to males ($c_m$)"
rownames(table_S1)[9]  <- "Male dispersal frequency ($x_m$)"
rownames(table_S1)[10] <- "Male dispersal frequency ($x_f$)"
rownames(table_S1)[12] <- "Number of patches ($k$)"
rownames(table_S1)[13] <- "Importance of local density ($\\psi$)"
rownames(table_S1)[14] <- "Male effect on density ($\\delta$)"
rownames(table_S1)[15] <- "Shape of density dependence ($\\alpha$)"
rownames(table_S1)[16] <- "Maximum fecundity ($r$)"

saveRDS(table_S1, file = "figures/tableS1.rds")
```


## Make Figures 3-4 and S1: effect of each parameter

```{r make_fig34}
find_percent_extinct <- function(variable, dataset, number_of_bins){
  
  dataset <- dataset %>% rename(value = !! variable)
  if(is.numeric(dataset$value) & length(unique(dataset$value)) > 5){
    dataset <- dataset %>% 
      mutate(value_bin = cut(value, number_of_bins),
             value = (gsub("]", "", gsub("[(]", "", value_bin))),
             value = map_dbl(strsplit(value, split = ","), ~ mean(as.numeric(.x)))) %>%
      group_by(value, value_bin) 
  } else {
    dataset <- dataset %>% group_by(value) 
  }
   
  output <- dataset %>%
    summarise(n_extinct = sum(went_extinct == 1),
              n_runs = n(),
              percent_extinct = 100 * n_extinct / n_runs,
              mean_extinction_time = mean(generation_extinct, na.rm = TRUE)) %>%
    mutate(parameter = variable,
           mean_extinction_time = replace(mean_extinction_time, is.nan(mean_extinction_time), NA))
  
  if(nrow(output) < 5) output$value <- as.character(output$value)
  output
}

make_one_plot <- function(dataset, response){
  
  dataset$y <- dataset %>% pull(which(names(dataset) == response))
  dataset$parameter <- nice_names$new[nice_names$original == dataset$parameter[1]]

  
  if(nrow(dataset) != 2){
    p <- ggplot(dataset, aes(value, y)) +
      geom_point(alpha = 0.6, colour = "#263056")
  } else {
    p <- ggplot(dataset, aes(value, y)) +
      geom_bar(stat = "identity", fill = "#ff6f61", width = 0.7, colour = "grey18")
  }
  
  if(dataset$parameter[1] %in% c("male_weighting", "density_dependence_shape")){
    p <- p + geom_vline(xintercept = 1, linetype = 2, colour = "grey18")
  }
  
  p + 
    stat_smooth(method = "loess", se = FALSE, colour = "#00a9e0") +
    labs(y = NULL, x = NULL) + 
    facet_wrap(~ parameter, labeller = label_parsed) + 
    theme_bw() + 
    theme(strip.background = element_rect(fill = "#fbd872"))
}

make_multiplot_data <- function(dat, cores){
  dat <- mclapply(variable_parameters, find_percent_extinct, dataset = dat, number_of_bins = 51, mc.cores = cores)
  out_list <- dat[order(sapply(dat, function(x) max(x$percent_extinct)), decreasing = TRUE)]
  keep <- which(sapply(out_list, nrow) > 1)
  out_list[keep]
}

make_multiplot <- function(multiplot_data, response){  
  
  left_axis_label <- ifelse(response == "percent_extinct", "% runs ending in extinction", "Mean generations until extinction")
  
  multiplot_data %>%
    map(~ make_one_plot(.x, response = response)) %>%
    grid.arrange(grobs = ., left = left_axis_label, bottom = "Parameter value")
}

if(!file.exists("figures/fig_S1.rds")){
  fig34_data <- make_multiplot_data(perfect_W_shredder, cores = 4)
  figS1_data <- make_multiplot_data(W_shredder, cores = 4)

  fig_3 <- make_multiplot(fig34_data, "percent_extinct")
  fig_4 <- make_multiplot(fig34_data, "mean_extinction_time")
  fig_S1 <- make_multiplot(figS1_data, "percent_extinct")

  fig_3 %>%  saveRDS(file = "figures/fig_3.rds")
  fig_4 %>%  saveRDS(file = "figures/fig_4.rds")
  fig_S1 %>% saveRDS(file = "figures/fig_S1.rds")
  fig_3 %>% ggsave(filename = "figures/fig_3.pdf", width = 9, height = 10)
  fig_4 %>% ggsave(filename = "figures/fig_4.pdf", width = 9, height = 10)
  fig_S1 %>% ggsave(filename = "figures/fig_S1.pdf", width = 9, height = 10)

} else {
  fig_3 <- readRDS("figures/fig_3.rds")
  fig_4 <- readRDS("figures/fig_4.rds")
  fig_S1 <- readRDS("figures/fig_S1.rds")
}
```

```{r fig.height=10, fig_width=14}
grid.newpage()
grid.draw(fig_3)
```
<br></br>
**Figure 3**: The plot shows the effect of each model parameter on the percentage of simulation runs in which the population went extinct, for simulations of a _Z_-linked _W_-shredder that completely prevents the production of daughters by females ($p_{shred} = 1$). The plot was generated by sampling evenly from the complete parameter space `r n_runs[[2,2]]` times using Latin hypercube sampling, such that each plot shows the marginal effect of one parameter while the other parameters vary independently across their possible ranges (shown by the relevant $x$-axis scale). Continuous parameters were grouped into 51 bins (each containing approximately 14,400 runs, i.e. ~2% of the total), and the lines were fitted using LOESS (locally estimated scatterplot smoothing). Note that the $y$-axis scales to the range of the data, and the panels are ordered by this range (showing the approximate relative importance of each parameter).

```{r fig.height=10, fig_width=14}
grid.newpage()
grid.draw(fig_4)
```
<br></br>
**Figure 4**: The plot shows the average number of generations until extinction, among the subset of runs in which extinction occurred. The density plot in the lower right shows the distribution of the response variable across all the parameter spaces in which the _Z\*_ allele was costly (defined as $c_f > 0.05$ and/or $c_m > 0.05$) or not costly ($c_f < 0.05$ and $c_m < 0.05$), and in which the _Z\*_ allele benefitted from strong gene drive in males (dp_{conv} > 0.95$) or not ($p_{conv} < 0.95$). Other details are the same as for Figure 3. 

```{r fig.height=10, fig_width=14}
grid.newpage()
grid.draw(fig_S1)
```
<br></br>
**Figure S1**: The plot shows the same information as in Figure 3. These data come from an independent set of `r n_runs[[1,2]]` simulation runs in which $p_{shred}$ was allowed to vary, in addition to all of the parameters shown in Figure 3.

### Make inset of Figure 4 - time to extinction frequency distribution

```{r make_inset}
median_gen <- perfect_W_shredder$generation_extinct %>% median(na.rm = T)
median_gen_no_cost <- perfect_W_shredder %>% filter(cost_Zdrive_female < 0.05 & cost_Zdrive_male < 0.05) %>%
  pull(generation_extinct) %>% median(na.rm=T)

annot <- data.frame(x = 51, y = 1.16, text = paste("Median:", median_gen))

inset <- perfect_W_shredder %>%
  mutate(no_cost = ifelse((cost_Zdrive_female > 0.05 | cost_Zdrive_male > 0.05) & Z_conversion_rate < 0.05, "No male drive, costly", NA)) %>%
  mutate(no_cost = replace(no_cost, (cost_Zdrive_female > 0.05 | cost_Zdrive_male > 0.05) &  Z_conversion_rate > 0.95, "Male drive, costly")) %>%
  mutate(no_cost = replace(no_cost, cost_Zdrive_female < 0.05 & cost_Zdrive_male < 0.05 & Z_conversion_rate < 0.05, "No male drive, not costly")) %>%
  mutate(no_cost = replace(no_cost, cost_Zdrive_female < 0.05 & cost_Zdrive_male < 0.05 & Z_conversion_rate > 0.95, "Male drive, not costly")) %>%
  filter(!is.na(no_cost) & !is.na(generation_extinct)) %>%
  mutate(no_cost = factor(no_cost, rev(c("No male drive, costly", "No male drive, not costly", "Male drive, costly", "Male drive, not costly")))) %>%
  ggplot(aes(generation_extinct, fill = no_cost)) +
  scale_x_log10(breaks = c(10,20,30,40,50,100,200,300,400,500,1000)) + 
  geom_density(alpha = 0.5) + 
  scale_fill_brewer(name = "Gene drive type", palette = "Set3") +
  annotation_logticks(sides="b") + 
  ylab("Density") + xlab("Generation in which extinction occurred") +
  theme_bw() + 
  theme(legend.position = c(0.8, 0.8))

inset %>% ggsave(filename = "figures/fig_4_inset.pdf", width = 4, height = 1.8)
```

